
---
title: "Toolkit - Getting started with R"
author: "Lauren Nelson"
date: "8/31/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(markdown)
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started with R 

## What you should already know:
* How to access the RStudio environment through Datahub
* How to run code (Run button or control/command + Enter)
* How to comment code (# or control/command + shift + C)

## What is included in this toolkit:
* Assigning values
  + Naming objects
  + Data types
* Creating objects using functions
  + Vectors
  + Matrices
  + Lists
* Calculations and Comparisons

## Assigning a value to an object

```{r}
#use "<-" as the operator to assign the value of 5 to the object named "x"
x <- 5
#return the value of x
x

#assign and return 
(x <- 5)

```


### Naming objects

* Names cannot start with numbers or symbols
* R is case sensitive!! 
* Best practices: 
  + Use lower case 
  + Use underscores to separate words in names 
  
```{r}
#some users prefer what is called 'camelCase' which uses a capital letter 
#to indicate a new word
camelCase <- "camel"
# CamelCase
camelCase

#we recommend using all lowercase and underscores to separate words
snake_case <- "snake"

```


### Data types

**Character values** 

* Indicated by quotation marks ("" or ''); best practice is to use double quotes 
* Can contain spaces, characters, symbols, and numbers

```{r}
#create a character value using double quotes
ch_double <- "dog"
#try using single quotes - this works, too
ch_single <- 'dog'

ch_double2 <- "turtle"

#the code below will not work, as R is looking for an object named "dog" 
#(since there are no quotations around it)
# ch_no <- dog

#if we create an objected named "dog"
dog <- "puppy"

#then re-run the line with no quotes, the value of "ch_no" will take on the
#value of the object "dog" which makes ch_no = "puppy"
ch_no <- dog
ch_no

#character values can be as long as you want
ch_long <- "this is a really really long string that i want to save"


```

**Numbers** can be stored in three ways. 

1. Numeric - both whole numbers or decimals
2. Integer - similar to whole number by indicated with an “L”
3. Complex 

```{r}
#numeric objects can be whole or decimal
num_whole <- 290
num_dec <- 290.9

#integers are indicated by adding an "L"
int <- 290L

#complex
complex <- 2+4i

```

**Logical:** Use all caps - TRUE or T, FALSE or F

```{r}
#two options for assigning a logical value to an object
logical <- TRUE
logical_1 <- T

#this does not save as a true logical value, rather it saves the string 
#"true" as a character
logical_lower <- "true"

```

## Creating objects using functions

**Vectors**

(Atomic) Vectors

- One dimensional, single data type
- Multiple ways to be created:
  + c() function
  + Using : operator for a vector of consecutive numbers
  
```{r}
#create a numeric vector using the c() function
vec_num <- c(1,5,6,94)
vec_num

#create a numeric vector using the : operator
vec_num2 <- 1:10

#create a character vector using the c() function
vec_char <- c("dog","cat","mouse")

#try creating a vector with multiple data types - this will force 290 to be
#stored as "290"
vec_multi <- c(290,"ph")

```


**Matrices**

* Multi-dimensional, single data type
* Created using matrix() function

```{r}
#create a matrix using the : operator to define the data included
matrix_1 <- matrix(data = 1:12, nrow = 3, ncol = 4, byrow = TRUE, dimnames = NULL)

```

**Lists**

* One-dimensional, multiple data types and/or objects
* Created using list() function
```{r}
#use the list function to see what happens if you add items of different types
my_list <- list(290, "290")

```

More on data frames next week!


### Using functions to describe objects 


A few examples:

* length() - how long is the object?
* class() - what type of object is it?
* typeof() - what data type is the object?

```{r}
#return information about matrix and vectors created above
length(matrix_1) 
class(matrix_1)
typeof(matrix_1)

length(vec_num2)
typeof(vec_num2)

typeof(vec_char)

```

## Calculations and comparisons

### Calculations

R can be used as a high-power calculator in the console and in the script. 

Calculations can be made on numbers and objects.

```{r}
#calculations can be performed on numbers
54*38743252349

#and objects
a <- 4
b <- 75

b/a

#calculations can be performed on vectors
vec_num2*10

```


New objects can be created as result of calculations.

```{r}
matrix_2 <- matrix_1 * 5

c <- b-a

```

**Using functions for calculations**

In addition to the operators above, there are many functions that can be used to do calculations. 

```{r}
#example: absolute value
abs(-90)

```

### Comparisons

Two values or objects can be compared to assess if equal (==), unequal (!=), less (< or <=), or greater (> or >=) and will return a true or false.

```{r}
5==40

"dog"=="cat"

a!=b

b>c

d <- (a+b)/c
d

d2 <- (a+b)>c
d2

```









