---
title: "Introduction"
author: "Lawrence Y. Tello and David Connell"
date: "8/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown Introduction

Welcome to PHW251: R for Public Health!

We're excited that you're here and look forward to a semester with you and R. The current file you are looking at is called an "R Markdown" or file, which has the extension ".Rmd". We will use this type of R file for the semester because of its readability and capacity for creating *neat* PDFs for you to turn in. 
  
\newpage

### RMarkdown Basics

Below is a R chunk, which is where you type your code. You can run a single line of code by holding down Ctrl (on Windows) or Command (on Mac) and hitting the Enter/Return key. If you want to run the entire chunk, you can click the green arrow on the right-hand side of the chunk. Try both of these methods for the code below!

```{r}
print("hello world")
```

```{r}
1+2
answer <- 1+2
```


```{r}
# this is a comment
# anything following a hash/pound sign within a R chunk is a comment. 
# comments are not evaluated when you run the code chunk
# for example, the following code will not be evaluated:

# 1+2

# but this code will:
1/2
```

Keyboard shortcut to knit: 
Alt + Shift + K (Windows)
Command + Shift + K (Mac)



