---
title: "Toolkit - Packages"
author: "Lauren Nelson"
date: "8/31/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Packages 

This toolkit will cover the following:

* What is a package?
* How do packages work in the R environment?
  + Library (Datahub)
  + Installation
  + Documentation
  + Loading
* Example

## What is a package?

* R packages are a collection of R functions, compiled code, and sample data
  + Some narrow, some broad
  + Some focused on specific industry needs, some focused on refining a specific type of task (import, stats, graphics, etc.)
  + Some meant to work with other packages (tidyverse is an example of this)
* Offer an expansion of the capabilities of base R
  + Improve efficiency or effectiveness of programming 
* Anyone can create a package 
* Thousands of packages available 
* RStudio does create and maintain several very helpful packages

## How to packages work in the R environment?

### Libraries

Packages are contained within a **library**.
* User library (user or cloud)
* System library (default packages)

To see all packages already stored in your environment:
```{r}
#see the paths to the library
.libPaths()

#uncomment and run code below to see packages that are available within 
#your environment
#open a new tab to see list
library()

#return list with more information to console; save to an object named "info"
info <- installed.packages()

#can also use the Packages tab to the right
```

There are a lot of packages already installed because we are using Datahub, which has an extensive list of packages available for student use.

### Installation 

Packages must be installed into your R Studio environment in order to be used
* Use install.packages() function
* Packages tab within RStudio
* CRAN (https://cran.r-project.org/web/packages/available_packages_by_date.html)
* Github
* Note: Often installing a new package will require the installation of other dependent packages (typically automatic)

```{r}
# install.packages("readr")
```

Package maintenance
* Remove
* Update
```{r}

#remove an unwanted package using the code below
# remove.packages("readr")

#return list of packages needing an update
old_list <- old.packages()

#update all packages
# update.packages("readr")
# update.packages()

#can also use the "update" feature on the Package tab

```

### Documentation

More information on packages
```{r}
#will open "help"
help(package = "readr")

```
Other resources:
[CRAN] (https://cran.r-project.org/web/packages/available_packages_by_name.html)
[R Documentation] (https://www.rdocumentation.org/)
[Tidyverse (when relevant)] (https://www.tidyverse.org/)


### Loading in session

To use any of the packages that are installed in your libraries, you must load them into your session using the library() function. Or you can use :: to utilize a single function of a package.
```{r}
#to have available for whole script - add to very top!
library(readr)

#option for a single use
# readr::read_delim...
```


### Example - so we want to use ggplot2...

As an example, let's say we want to use a functionality of the package ggplot2. 

1. See if ggplot2 is installed in our environment

2. Make sure to understand the documentation
[Tidyverse] (https://ggplot2.tidyverse.org/) 
[R Documentation] (https://www.rdocumentation.org/packages/ggplot2/versions/3.3.2)
[CRAN site] (https://cran.r-project.org/web/packages/ggplot2/index.html)
[CRAN documentation] (https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf)

3. Load ggplot2 into our session

```{r}
#start script 
library(ggplot2)

ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()

#or

# ggplot2::ggplot(mpg, aes(displ, hwy, colour = class)) + 
  # geom_point()


```





