---
title: "Epidemic Curves"
author: "Manasa Susarla"
date: '2022-11-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome!

In this module, we will explore how to create epidemic curves with the incidence2 package.

After this module, students should be able to:

- Create basic and stratified epidemic curves

- Modify epidemic curves to represent different intervals


```{r load library, message = FALSE, warning = FALSE}
library(tidyverse)
library(magrittr)
library(pacman)
pacman::p_load(rio)
pacman::p_load(incidence2)

```

We will work with simulated Ebola epidemic data.

```{r load data}

Ebola_data <- import("https://github.com/appliedepi/epirhandbook_eng/raw/master/data/case_linelists/linelist_cleaned.xlsx")

head(Ebola_data)

```

Let's move on to plotting the epidemic curve.

```{r}
##First make sure that your date column is in Date Format
Ebola_data$date_onset <- as.Date(Ebola_data$date_onset)

#Step 1: Create an incidence object

epi_day <- incidence(x = Ebola_data, date_index = date_onset, interval = "day")

#Step 2: Plot

plot(epi_day)

```

We can modify how we want to aggregate our data using the "interval" argument  

```{r}
epi_month <- incidence(x = Ebola_data, date_index = date_onset, interval = "month")

plot(epi_month)

```

We can also group by a particular category.

```{r}
age_group <- incidence(x = Ebola_data, date_index = date_onset, interval = "week", groups = age_cat, na_as_group = TRUE)

plot(age_group, fill = age_cat)+
labs(fill = "Age Category")

```

AGGREGATED COUNTS

Let's create a dataset of weekly aggregated counts by hospital.

```{r}
count_data <- Ebola_data %>% 
  group_by(hospital, date_hospitalisation) %>% 
  summarize(n_cases = n()) %>% 
  ungroup()

head(count_data)

```

Plot the epidemic curve. 

```{r}
epi_counts <- incidence(count_data, date_index = date_hospitalisation, count = n_cases, interval = "week", groups = hospital)


plot(epi_counts, fill = hospital)  

```

