---
title: "Purrr and functional programming"
author: "Will Wheeler"
date: "11/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(dplyr)
library(magrittr)
library(purrr)
library(readxl)


```


### Why use purrr and functional programming?

Generally if you can write code that takes up fewer lines, reduces complexity, and follows logical progression, it makes it easier to understand and reduces likelihood of errors. 

Purrr is RStudio's package for functional programming and it contains a lot of powerful tools ([here's the link to the documentation](https://purrr.tidyverse.org/)). For anyone interested in advanced programming techniques using purr and why functional programming is useful, I highly recommend watching [this video](https://www.youtube.com/watch?v=bzUmK0Y07ck) and [reading this chapter in R for Data Science](https://r4ds.had.co.nz/iteration.html).

We talk a lot about the rule of three - if you have to do a thing more than three times, write a function!  Using functions make your code less susceptible to random error.  

If I have to run the same process on three different data frames, if I write out code three times to handle each data frame, then if I make a change, I have to make sure I make the exact same change on each code block.  If I don't, I've introduced error.  And that kind of error can often be really hard to find.

And if I can code in a way that just handles whatever file is sent my way, I have to make even fewer changes and am less susceptible to random error. (Note - you always of course have to watch out for systemic errors in your code, but that's why we include QA on datasets for analyses!)


## Examples

### Loading multiple spreadsheet sheets

```{r}

# Long way around
site_file <- "data/test_sites.xlsx"

siteA <- read_excel(site_file, sheet = "site_A") %>%
  select(uniqueid, age, gender, payer, undetectable, referred)

siteB <- read_excel(site_file, sheet = "site_B") %>%
  select(uniqueid, age, gender, payer, undetectable, referred)

siteC <- read_excel(site_file, sheet = "site_C") %>%
  select(uniqueid, age, gender, payer, undetectable, referred)

all_sites0 <- bind_rows(siteA, siteB, siteC)

```

Let's say I want to add a previous_visit data point, but I only remember to do it on two of the datasets

```{r}

# Long way around
site_file <- "data/test_sites.xlsx"

siteA <- read_excel(site_file, sheet = "site_A") %>%
  select(uniqueid, age, gender, payer, undetectable, previous_visit, referred)

siteB <- read_excel(site_file, sheet = "site_B") %>%
  select(uniqueid, age, gender, payer, undetectable, previous_visit, referred)

siteC <- read_excel(site_file, sheet = "site_C") %>%
  select(uniqueid, age, gender, payer, undetectable, referred)

all_sites0 <- bind_rows(siteA, siteB, siteC)

```

This is just an example, but if I don't notice that, I could be in big trouble.

# A better way around

```{r purrr_way}
import_sheets <- c("site_A", "site_B", "site_C")


all_sites3 <- map_df(import_sheets, 
                           ~read_excel(site_file, sheet = .x) %>%
                            select(uniqueid, age, gender, payer, previous_visit, undetectable, referred)
)

# Even a little fancier - excel_sheets() function will pull in all sheets in a file, so you don't even have to create import sheets!
# Note that they don't even necessarily have the same headers - you can use the recode() function to rename columns into standard names

all_sites4 <- map_df(excel_sheets("data/test_sites.xlsx"),
                     ~read_excel(site_file, sheet = .x) %>%
                       select(uniqueid, age, gender, payer, previous_visit, undetectable, referred) %>%
                       mutate(age_cat = case_when(age %in% c(0:24) ~ "< 25",
                                                  age %in% c(25:49) ~ "25-49",
                                                  age %in% c(50:150) ~ "> 50"))
)



```

## How about another example that might help in an epi/biostats milieu?

Let's say I wanted to run a model on a number of datasets

```{r}

# create three distinct data frames one for each unique value of the cyl column - results in a list of dataframes

test <- mtcars %>%
  split(.$cyl)

# generate a linear model of mpg as a function of wt for each cyl data frames

test <- mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ wt, data = .x)) 

# from that resulting list, pull out the model summary for each

test <- mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ wt, data = .x)) %>%
  map(summary) 

# from that resulting list, pull out only the r.squared values (you could pull out whatever you need to evaluate)

test <- mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ wt, data = .x)) %>%
  map(summary) %>%
  map_dbl("r.squared")

test
```

