---
title: "tableone"
author: "Ilya Moskalenko"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
# PREPARING DATA
```{r installing and loading packages, include=FALSE}
#install.packages("tableone")
library(tableone)

#install.packages("dplyr")
library(dplyr)

#install.packages("nhanesA")
library(nhanesA)
```

```{r reading in and viewing the data}
# the nhanes() function reads in NHANES specific data ("DEMO_H" in this case)
data <- nhanes("DEMO_H")

# viewing the first 6 rows of the data set
head(data)
```

We see that there are 47 variables in the NHANES dataframe that we read in. Keeping the best practices of data science in mind, the **first step** would be to want select the **columns** that we are interested in analyzing. 

1. **Select** variables that we are interested in:
    * Gender (RIAGENDR)
    * Age (RIDAGEYR)
    * Race (RIDRETH3)
    * Education (DMDEDUC2)

```{r selecting variables of interest}
data <- data %>% select(
               c("RIAGENDR", # Gender 
                 "RIDAGEYR", # Age
                 "RIDRETH3", # Race
                 "DMDEDUC2")) # Education
               
head(data)
```

2. **Translate** variable names and data points into meaningful data. The NHANES data set is a big data set and the information is coded as factors (the numbers we see in the raw dataset) because they take up less space. The *translation* process will convert the numbers into their corresponding real values.

```{r translating the NHANES code into meaningful data}
data <- nhanesTranslate("DEMO_H",
                 c("RIAGENDR", # Gender 
                   "RIDAGEYR", # Age
                   "RIDRETH3", # Race
                   "DMDEDUC2"), # Education
                   data = data)

head(data)
```

3. *Rename* column names to something more legible after performing the translation.

```{r changing the column names}
data <- data %>% rename( 
       Gender= RIAGENDR,
       Age= RIDAGEYR,
       Race= RIDRETH3,
       Education= DMDEDUC2)

head(data)
```

\newpage

# USING TABLEONE

Create simple **Table 1** using **tableone** package.

```{r creating our simple table one}
tableone <- CreateTableOne(data = data, # the data set
               vars = c("Gender", "Age", "Race", "Education"), # all variables
               factorVars = c("Gender", "Race", "Education")) # factors only
              
tableone
```

\newpage

## showAllLevels and cramVars

### showAllLevels

We see that the **Gender** variable only shows the percent of the study population that is “Female”. To show both “Female” and “Male” categories, we utilize the **showAllLevels = TRUE** argument within a **print()** function. This will also show all the factor levels for all of the variables in the dataframe.

```{r showing all levels}
print(tableone, showAllLevels = TRUE)
```

### cramVars

If we want to better utilize the space of a certain categorical variable (factor) we can print it all on one line using the **cramVars = ** argument as done with gender in this instance. NOTE: This only works with *binary classifiers* because otherwise it would get too crammed to discern the data.

```{r cramming certain levels on one line}
print(tableone, cramVars = "Gender")
```
\newpage

## Addressing Data that is NOT Normally Distributed

If for instance, some of our continuous data is not normally distributed it would be more informative to report a median with an IQR rather than just a percent composition value to a journal publication. You can do this by indicating which variables are not normally distributed. NOTE: For the purposes of this tutorial, we are just showing how to use this feature on the **Age** variable without evaluating the normality of the distribution.

```{r}
print(tableone, nonnormal = "Age")
```

\newpage

## Showing Only Categorical or Only Continuous Variables

### Only Categorical

```{r}
# Categorical variables only
print(tableone$CatTable)
```
### Only Continuous

```{r}
# Continuous variables only
print(tableone$ContTable, nonnormal = "Age")
```

\newpage

## Stratifying Your Data

Stratifying our data across a variable such as **Gender** is potentially the most useful part of using the **tableone** package to analyze study data. For the purposes of this tutorial, the data was stratified based on *Gender* for demonstration purposes. However, data will typically be stratified by the exposure of interest that *Epidemiologists* and *Biostatisticians* are trying to evaluate. This is especially useful for studies that have multiple treatment arms (control, treatment 1, treatment 2, treatment 1 + treatment 2).

Stratifying your data also yields **p-values**. The *chisq.test()* is the default test for **categorical variables**. The *oneway.test() (regular ANOVA)* is the default test for **normal continuous variables**. Meanwhile, the The *kruskal.test()* is the default test for **nonnormal continuous variables**

### Stratified Table 1 with Normal Values
```{r}
tableone_stratified <- CreateTableOne(data = data, # the data set
               vars = c("Age", "Race", "Education"), # all variables
               factorVars = c("Race", "Education"), # factors only
               strata = "Gender") # stratifying variable (exposure of interest)
              
print(tableone_stratified)
```

### Stratified Table 1 with Nonnormal Values
```{r}
print(tableone_stratified, 
      nonnormal = "Age")
```

\newpage

# Downloading Your Table 1 to your Machine

This code prepares our created *Table 1* for downloading.
```{r}
tableone_csv <- print(tableone_stratified,
                 printToggle = FALSE) # need to include this argument
tableone_csv
```

This code creates a csv file in your working directory.
```{r}
write.csv(tableone_csv, # name of your Table 1
          file = "Table_1.csv") # name of the created file
```

