---
title: "Week 7: Correcting for unexpected or invalid values"
author: "Lauren Nelson"
date: "9/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
```

**Example**
Testing counts by group

Goal: Calculate positivity rates by day for Group 1 and Group 2 in August 2020

Columns:

* test_date
* group
* total_tests
* pos_tests
* group_pop


Import & open table 
```{r}

file_path <- "data/testing_example.xlsx"

testing <- readxl::read_excel(
  file_path,
  na="NA"
)

#confirm total_tests is now numeric
str(testing)

```


## CORRECTIONS

1. Test dates - out of range and missing
2. Group - invalid names, remove missing and group 3, change groups to city names
3. Total Test - outlier; "NA" instead NA; numeric; change missing to 0
4. Positive tests - instance where > than total_tests; set missing to 0


**Test dates** 

* out of range
* missing 
```{r}
#Method 1: dplyr::filter and drop_na
testing1a <- testing %>%
  drop_na(test_date) %>%
  filter(test_date >= as_date("2020-08-01") & test_date <= as_date("2020-08-31"))
 
#Methood 2: filter 
testing1b <- testing %>%
  filter(test_date >= as_date("2020-08-01") & test_date <= as_date("2020-08-31") & !is.na(test_date))

```

**Group**

* Invalid names
* Remove missing and group 3
* Change groups to city names
```{r}
#Method 1: if_else/case_when and recode
groups <- testing %>%
  select(group) %>%
  distinct() %>%
  pull()

groups

testing2a <- testing1a %>%
  mutate(group = case_when(
    group %in% c("group 1","1","gp one","gp 1") ~ "G1",
    group %in% c("group 2","Gp Two") ~ "G2",
    TRUE ~ NA_character_
  )) %>%
  drop_na(group) %>%
  ##mutate(group = recode(group, "G1"="San Diego", "G2"="Los Angeles"))
  mutate(group = recode(group, `G1`="San Diego", `G2`="Los Angeles"))

testing2a



## Sn50
## - `Grp1` can be replaced with "Grp1".  R handle 3 types of quotes?  " ' and `
#Method 2: if_else/case_when
testing2b <- testing1a %>%
   mutate(group = case_when(
    group %in% c("group 1","1","gp one","gp 1") ~ "San Diego",
    group %in% c("group 2","Gp Two") ~ "Los Angeles",
    TRUE ~ NA_character_
  )) %>%
  drop_na(group) 




#Method 3: if_else/case_when & grepl
testing2c <- testing1a %>%
   mutate(group_cat = case_when(
    grepl("1|one",group, ignore.case=TRUE) ~ "San Diego",
    grepl("2|two",group, ignore.case=TRUE) ~ "Los Angeles",
    TRUE ~ NA_character_
  )) %>%
  drop_na(group_cat)

#confirm categories were created correctly
table(testing2c$group_cat,testing2c$group)

```

**Total Test**

* "NA" instead NA
* confirm numeric
* outlier
* change missing values to 0
```{r}




#Method 1: Handled at import
testing <- readxl::read_excel(
  file_path,
  na="NA"
)

#Method 2: Handled in mutate
testing3 <- readxl::read_excel(
  file_path
)

testing3a <- testing3 %>%
  mutate(total_tests = na_if(total_tests,"NA"),
         total_tests = as.numeric(total_tests))
 
#outlier
#Method 1: mutate, if_else using row_number or other value
testing3b <- testing2a %>%
  mutate(total_tests = if_else(row_number()==23, 450, total_tests)) %>%
  replace_na(list(total_tests=0))
```

**Positive tests**

* greater than total tests to missing
* change missing to 0
```{r}
#Method 1: replace_na
testing4a <- testing3b %>%
  mutate(pos_tests = if_else(pos_tests>total_tests, NA_real_, pos_tests)) %>%
  replace_na(list(pos_tests=0, total_tests=0))

#Method 2: single if_else
testing4b <- testing3b %>%
  mutate(pos_tests2 = if_else(pos_tests>total_tests | is.na(pos_tests), 0, pos_tests)) 
         

```

**Percent Positive**

Calculate percent positive & clean up if needed
```{r}

testing5a <- testing4a %>%
  mutate(pct_pos = round(100*pos_tests/total_tests,0))


#clean up values that are NaN - set to 0
testing_clean <- testing5a %>%
  mutate(pct_pos = if_else(is.nan(pct_pos),0,pct_pos))

```


