---
title: "Week 7: Detecting unexpected or invalid values"
author: "Lauren Nelson"
date: "9/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
```

**Example**
Testing counts by group

Goal: Calculate positivity rates by day for Group 1 and Group 2 in August 2020

Columns:

* test_date
* group
* total_tests
* pos_tests
* group_pop


**Old-fashioned way**

Import & open table 
```{r}

file_path <- "data/testing_example.xlsx"

testing <- readxl::read_excel(
  file_path
)

#fix the NA issue with the total_tests column
testing <- readxl::read_excel(
  file_path,
  na="NA"
)

#confirm total_tests is now numeric
str(testing)


## gt = greater than
#create new logical variables to help with looking at our data
#total_gt_pos will return TRUE where total_tests>pos_tests
#total_gt_pop will return TRUE where group_pop>total_tests
testing_temp <- testing %>%
  mutate(total_gt_pos = total_tests>pos_tests,    ##### Pff.... she looked at col with FALSE as result!  more logical to be less_than, sritct, no <= .   # video ~8:03   (or i guess she is looking for TRUE to say data pass validation test, those that are FALSE fail validation)
         total_gt_pop = group_pop>total_tests)


## testing_temp 
#one instance where pos_tests > total_tests (data entry error?)
#one instance where total_tests > group_pop (outlier?)

```

**Utilize descriptive functions**

* Unique - list of unique values
* table - frequency
  + useNA to include frequency of NA ("ifany" to include when relevant, "always" to include regardless)
  + can cross tab
* summary - numeric stats
* sum(is.na()) - number missing
```{r}
# test_date
table(testing$test_date)   ## Sn50


table(testing$test_date, useNA="always")
summary(as_date(testing$test_date))
sum(is.na(testing$test_date))

#need to limit dates to aug 2020 only; get rid of missing

# group
unique(testing$group)
table(testing$group, useNA = "ifany")
table(testing$group, useNA = "always")   ##
table(testing$group )

#need to standardize the group name; remove group 3 and NA's before analyzing

# total_tests
summary(testing$total_tests)
sum(testing$total_tests)        ## Sn50
sum(testing$total_tests, na.rm=T)        ## Sn50

sum(is.na(testing$total_tests))

#need to look into the outlier - 45,000 tests; handle missing values

# pos_tests
summary(testing$pos_tests)
sum(is.na(testing$pos_tests))

#okay, but needs comparison with total tests

# group_pop
table(testing$group, testing$group_pop, useNA="always")

table(testing$group, useNA="always")
table(testing$group_pop, useNA="always")



#okay for group 1 and 2


```

**Plots**

ggplot() - flexible plotting for one or two variables
```{r}

#one variable
#categorical - group
ggplot(data = testing, mapping = aes(x = group)) +
  geom_bar()

table(testing$pos_tests)  ### Sn50
table(testing$group, useNA="always")  ### Sn50, text equiv of ggplot above


#continuous - positive tests
ggplot(data = testing,mapping = aes(x = pos_tests)) +
  geom_histogram(binwidth=5)

table(testing$pos_tests, useNA="always")  ### Sn50, text equiv of ggplot above, but need to use some group_by a/o case_when to have mul values binned to same column


#continuous - total tests
ggplot(data = testing,mapping = aes(x = total_tests)) +
  geom_histogram(binwidth = 50) +
  xlim(0,500)


# need to use the xlim to remove the outlier from the plot
  
```


```{r}

#two variables
#two categorical (continuous can be used here as well)
ggplot(data = testing, mapping = aes(x = total_tests, y = pos_tests)) + 
  geom_count() +
  xlim(0,500)


## Sn50
table( testing$total_tests, testing$pos_tests )

## Sn50
ggplot(data = testing, mapping = aes(x = total_tests, y = pos_tests)) + 
  geom_point() +
  xlim(0,500)

ggplot(data = testing, mapping = aes(x = total_tests, y = group)) + 
  geom_point() +
  xlim(0,500)


#one categorical, one continuous 
#total_tests by group
ggplot(data = testing, mapping = aes(x = total_tests, y = group)) + 
  geom_boxplot() +
  xlim(0,500)

#two continuous
#total_tests vs pos_tests
ggplot(data = testing, mapping = aes(x = total_tests, y = pos_tests)) + 
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  xlim(0,10) +
  ylim(0,10)

#total_tests vs group_pop
ggplot(data = testing_temp, mapping = aes(x = group_pop, y = total_tests)) + 
  geom_point()  +
  geom_abline(intercept = 0, slope = 1) 

```








