---
title: "Week 7 - Missing data"
author: "Lauren Nelson"
date: "9/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Missing data 

* Types
* Properties

**NA**

Not Available - Logical constant indicating a missing value. Can be coerced to any other data type vector: NA_integer_, NA_real_, NA_complex_, NA_character_ 
```{r}

#NA
a <- NA
a
class(a)
length(a)

#Expressions with NA always return NA
a+1
a>1
# a==NA
is.na(a)

#another method for determining if values are the same or "identical"
identical(a,NA)

identical(5,5)

#NA of other data types - commonly needed in if_else() and case_when()
a_char <- NA_character_
class(a_char)
is.na(a_char)

a_num <- NA_real_
class(a_num)
is.na(a_num)


#NA vs "NA" - not the same!
c <- "NA"
c
class(c)
is.na(c)
identical(a,c)

```

**NULL**

Object returned when an expression or function results in an undefined value. Really an empty value. 
```{r}
#NULL
b <- NULL
b
class(b)
length(b)
b+1
is.null(b)


#NULL vs "NULL" - not the same
d <- "NULL"
d
is.null(d)
identical(b,d)


#NULL vs NA - also not the same!
a <- NA
b <- NULL

is.null(a)
is.na(b)
identical(a,b)



```

**NaN**

Not A Number - logical vector of a length 1 and applies to numerical values only (not integer)
```{r}
#generate a NaN value
e <- sqrt(-1)
e

e+1

is.nan(e)

#is.na will pick up NaN values
is.na(e)

#is.nan will not pick up NA values
is.nan(a)


```

**Infinite**

Large number or a product that is a result of division by zero. 
```{r}
#create an infinite value
f <- 10/0
f

-10/0

is.infinite(f)

#infinite are not recognized as NA
is.na(f)

```


## Assessing missingness

* Finding NA, NULL, etc.
* Counting NA, NULL, NaN, etc.

```{r}

#vector with NA, NULL, and NaN
v <- c(NA, 3, 5, NaN, NULL, NA, NaN, 6, 7)
#print vector an notice that NULL does not appear in vector; 
#it is not recognized as a value in the vector
v
length(v)
v[5] #returns NA, not NULL

#calculate mean of vector
mean(v)

#remove NA's from mean calculation
mean(v, na.rm=TRUE)

#return T/F for each element in the vector
is.na(v)

#return number of NA and NaN elements
sum(is.na(v))
sum(is.nan(v))

#return indexes of NA values
which(is.na(v))

#create a new vector of non-NA values
z <- v[!is.na(v)]
z

mean(z)

```


