---
title: "wk5_dplyr_manipulating_columns"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) #dplyr & tibble are included in this

```

## Tidying (or manipulating) table columns
* Extracting columns
  + select(), pull()
* Rename
* Adding
  + mutate()


```{r}

#import file of infectious disease data by county
file_path <- "https://data.chhs.ca.gov/dataset/03e61434-7db8-4a53-a3e2-1d4d36d6848d/resource/75019f89-b349-4d5e-825d-8b5960fc028c/download/idb_odp_2001-2018.csv"

#import as is; not going to fix the NA values in Rate (will address later)
ids <- read_csv(
  file_path
)

#check out table structure/types
str(ids)

```

**select()**

* select multiple columns to extract/subset
* : to select range
* c() to specify list of columns/variables (use - to exclude)
* ! to take the complement of a set of columns
* helper functions
  + everything(), last_col()
  + pattern matching: starts_with(), ends_with(), contains(), matches()

```{r}
#drop rate (use - or !)
id_sub <- select(ids, !Rate)
id_sub <- select(ids, -Rate)

#drop lower and upper ci
id_sub <- select(ids,-contains("CI"))
#or
id_sub <- select(ids,!contains("CI"))
#or use ends with
id_sub <- select(ids,!ends_with("CI"))

#create new table that excludes the confidence intervals and the pre-calculated rate
id_sub <- select(ids,c(Disease,County,Year,Sex,Cases,Population))
#or
id_sub <- select(ids,Disease:Population)
#select and change order of columns
id_sub <- select(ids,c(Population,Disease,County,Year,Sex,Cases))



```

**pull()**

create a vector from column values

```{r}
#create a vector of unique diseases in table
diseases <- id_sub %>%
  distinct(Disease) %>%
  pull(Disease)

#since there is only one column, could also use 
diseases <- id_sub %>%
  distinct(Disease) %>%
  pull()

#named vector of case counts, named by county
county_counts <- id_sub %>%
  filter(Disease=="Zika Virus Infection" & Year==2018 & Cases>0 & Sex=="TOTAL") %>%
  pull(Cases,name=County)

county_counts

```

**rename()**

* provide new name = old name
* rename_with() can be used to apply a function to all column names
```{r}

#rename single variable (and keep all others the same)
ids_rename <- rename(ids, disease=Disease)

#rename multiple variables
ids_rename <- rename(ids, disease=Disease, county=County)

#rename all - this line will rename all variables as lower and replace spaces with _ 
#the .x is used as a placeholder for column names since this will be applied to all columns
ids_rename <- rename_with(ids, ~ tolower(gsub(" ","_",.x,fixed=TRUE)))

#can also rename as part of the select statement, but will only include selected columns
ids_rename2 <- select(ids,disease=Disease)
```

**mutate()**

* create new columns, or overwrite existing ones
* values
  + vector of length 1
  + vector of length of the data frame
  + NULL to remove column
  + calculation or function based one existing columns (a lot of options!!)
```{r}

#create a table with Campylobacteriosis for LA; only keep rows with TOTAL
id_campy <- ids_rename %>%
  filter(disease=="Campylobacteriosis" & sex=="TOTAL") %>%
  arrange(year)

#use mutate to clean up the Rate column
ids_campy_rate <- id_campy %>%
  mutate(rate = na_if(rate,"-"),
         rate_num = as.numeric(rate),
         rate_clean = str_remove(rate,"[*]"),
         rate_clean = as.numeric(rate_clean),
         rate_calc = 100000*cases/population, 
         rate_check = near(rate_clean,rate_calc,tol=0.001),
         high_rate = rate_calc>10
  )

#clean up other columns
ids_campy_clean <- ids_campy_rate %>%
  select(-c(sex,rate,rate_num,rate_clean,rate_check,high_rate)) %>%
  mutate(county=str_to_title(county),
         state = "CA",
         county_small = if_else(population<100000,"small","not small"),
         county_cat = case_when(
           population<100000 ~ "small",
           population<1000000 ~ "med",
           TRUE ~ "large"
         ),
         ci = paste0("(",`lower_95%_ci`,", ",`upper_95%_ci`,")")
         ) 

#subset for LA and calculate difference between years
id_campy_la <- ids_campy_clean %>%
  filter(county=="Los Angeles") %>%
  select(disease,year,cases) %>%
  mutate(cumulative_cases = cumsum(cases),
         cumulative_mean_cases = cummean(cases),
         cases_last_year = lag(cases),
         cases_change = cases - cases_last_year
         )



```