---
title: "wk5_dplyr_manipulating_rows"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) #readr, dplyr, and tibble are included in this
```

## Tidying Rows
* Extracting rows
  + filter()
  + slice() and more
  + sample_n & sample_frac()
  + distinct()
* Adding - add_row()
* Order - arrange()



```{r}

#import table of infectious disease by county in CA
file_path <- "https://data.chhs.ca.gov/dataset/03e61434-7db8-4a53-a3e2-1d4d36d6848d/resource/75019f89-b349-4d5e-825d-8b5960fc028c/download/idb_odp_2001-2018.csv"

#try out read_csv with no specifications
ids <- read_csv(
  file_path
)

ids <- read_csv(
  file_path,
  na = c("", "NA","-")
)

#check out structure 
str(ids)



```

**filter()**

Extract (or subset) rows from table based on column/variable values

* Key operators/functions
  + equal (==), not equal (!=), greater than (> or >=), or less than (< or <=)
  + %in% or !%in% (not in)
  + is.na() to return missing; !is.na() to return non-missing
  + between(), near()
* Multiple expressions can be combined using & (and) or | (or)
```{r}
#filter to have 
# Salmonellosis, Listeriosis
# Total 
# Bay Area
# Population >100000
# 2018

#filter where disease is Salmonellosis or Listeriosis
id_sub <- filter(ids, Disease %in% c("Salmonellosis","Listeriosis"))
#or
id_sub <- filter(ids, Disease=="Salmonellosis" | Disease=="Listeriosis")

#filter where disease is Salmonellosis or Listeriosis and total sex (don't need male or female)
id_sub <- filter(ids, Disease %in% c("Salmonellosis","Listeriosis") & Sex=="TOTAL" )
id_sub <- filter(ids, Disease %in% c("Salmonellosis","Listeriosis"), Sex=="TOTAL" )

#filter where disease is Salmonellosis or Listeriosis, total sex, bay area
bay_area <- c("ALAMEDA","CONTRA COSTA","SAN FRANCISCO", "MARIN", "SAN MATEO", "SONOMA", 
              "NAPA", "SOLANO","SANTA CLARA")

id_sub <- filter(ids, Disease %in% c("Salmonellosis","Listeriosis") & 
                   Sex=="TOTAL" & County %in% bay_area)

id_sub <- filter(ids, Disease %in% c("Salmonellosis","Listeriosis") & 
                   Sex=="TOTAL" & County %in% bay_area & Year==2018)

#other examples
#keep rows that contain the maximum year
id_sub_a <- filter(ids, Year==max(Year) )

#keep rows from 2017-2018 
id_sub_b <- filter(ids, between(Year,2017,2018) )

#keep rows that are not missing rate
id_sub_c <- filter(ids, !is.na(Rate))

```

**slice()**

* slice(): use : or c()
* slice_head() or _tail(): n for number of rows
* slice_min() or _max(): 
  + n for number of rows
  + order_by to select column to sort by
  + with_ties to indicate if tie rows are included
```{r}
#from data for all counties, for 2018, Listeriosis
lister_2018 <- filter(ids,Year==2018 & Disease=="Listeriosis" & Sex=="TOTAL")

#slice using : 
lister_2018a <- slice(lister_2018, 15:n())

#slice using c()
lister_2018b <- slice(lister_2018, c(5,10,15,20,25))

#slice_head
lister_head <- slice_head(lister_2018,n=10)

#slice_tail
lister_tail <- slice_tail(lister_2018,n=10)

#slice_max
lister_max <- slice_max(lister_2018, order_by=Cases, n=10, with_ties = FALSE)

#slice_min
lister_min <- slice_min(lister_2018, order_by=Cases, n=3, with_ties = TRUE)


```

**sample - sample_n & sample_frac()**

* size: sample_n use number of rows; sample_frac using proportion of rows (between 0 and 1)
* replace: TRUE will sample with replacement; FALSE will not sample with replacement
```{r}

#random sample of 10 counties
lister_samp_n <- sample_n(lister_2018,size=10,replace=FALSE)

#random sample of 10% of counties
lister_samp_frac <- sample_frac(lister_2018,size=.1,replace=FALSE)

```

**distinct()**

* single or multiple columns
* keep_all to keep columns not listed (TRUE) or exclude columns not listed (FALSE)
```{r}
#return distinct rows
id_distinct <- distinct(id_sub)

#return unique counties
id_distinct <- distinct(id_sub, County)

#return unique county and disease combinations; keep all columns
id_distinct <- distinct(id_sub, County, .keep_all=TRUE)

#return unique county and disease combinations
id_distinct <- distinct(id_sub, County, Disease)

```


**tibble::add_row()**

* provide name/value pairs for columns that already exist
* unset columns will get an NA value
```{r}
#add row to id_distinct with Disease and County specified
#add another row with only County specified
id_add <- id_distinct %>%
  add_row(Disease = "E. Coli", County = "San Diego") %>%
  add_row(County = "Orange")


```


**arrange()**

* indicate column(s) to sort by
* use desc() to indicate a descending sort
```{r}
#order by Disease
id_order <- arrange(id_sub, Disease)

#order by cases 
id_order <- arrange(id_sub, Disease, Cases)

#order by cases descending
id_order <- arrange(id_sub, Disease, desc(Cases))

#order by cases descending
id_order <- arrange(id_sub, desc(Disease), desc(Cases))
#or
id_order <- arrange(id_sub, desc(Disease,Cases))


```

