---
title: "Toolkit - Vectors & Lists"
author: "Lauren Nelson"
date: "9/14/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vectors & Lists 
- Creating
- Indexing/extracting
- Modifying 


**Vectors**
One dimensional objects that contain a single data type: numeric, character, complex, logical

- Creating vectors
  + c() function
  + Using : operator for a vector of consecutive numbers
  + seq()
  + from data frame/tibble columns
```{r}
#create vector using c()
vec_num <- c(1,5,6,94,NA)
vec_num

vec_char <- c("dog","cat","mouse")
vec_char

#utilize previously learned functions to see info about vectors
length(vec_char)
str(vec_char)
typeof(vec_char)

#create vector using :
vec_num1 <- 1:10
vec_num1

#create vector using seq()
vec_seq <- seq(20,25,by=0.25)
vec_seq

vec_seq2 <- seq(20,25,length.out=10)
vec_seq2

typeof(vec_seq2)

#what happens when different data types are included? 
vec_multi <- c(1,2,3,"dog")
vec_multi

str(vec_multi)

#create a vector from column in data frame 
testing_plain <- data.frame(
  county = c("Alameda","Contra Costa","Marin"),
  total_tests = c(500,745,832),
  pos_tests = c(43, 32, 30)
)

v_county <- as.character(testing_plain$county)
v_county

#insert tangent covered later in the video
#create a vector of counties we are interested in
county_list <- c("San Diego","Marin")

#compare our vector with the list of counties to see which are in the list
v_county %in% county_list

#named vectors
named_v <- c("day1" = "Monday", "day2" = "Tuesday", "day3" = "Wednesday" )
named_v

```

- Indexing a vector
  + integers (positive or negative)
  + logical vectors
  + character vectors/names
  + which()
```{r}
#create vector for use below
new_vec <- seq(0,50,by=5)

#5th element
new_vec[5]
new_vec

#elements 5-10
new_vec[5:10]

#elements 3,6,9
new_vec[c(3,6,9)]

#all elements except 3,6,9
new_vec[c(-3,-6,-9)]

#all elements except 5,6,7
new_vec[-5:-7]

#shorter vector
short_vec <- new_vec[5:10]
short_vec

#logical vector index
short_vec[c(FALSE,TRUE,FALSE,TRUE,FALSE,FALSE)]

#by condition
short_vec[short_vec>30]

#which
short_vec

#returns position of values that match the condition
which(short_vec>30)

#returns elements that match the condition
short_vec[which(short_vec>30)]

#see names of elements
named_v
names(named_v)

#index by name
named_v["day1"]

named_v[c("day1","day2")]


```

- Modify vectors
  + Change element
  + Delete 
  + Combine vectors
  + Calculate
```{r}
#create multiple vectors to use below
v1 <- c(2,7,3,21,98)
v2 <- 1:5
v3 <- c("a","b","c","d","e")
v4 <- c(1,3)

#change an element
v1[5] <- 94
v1

#add an element
v1[6]<- NA
v1

#delete an element
v1[-6]
#or
v1[1:5]

#clear vector values
#create vector to delete
test <- v1

#clear values from vector
test <- NULL

#combine vectors 
#same type
v1_2 <- c(v1,v2)
v1_2

#numeric and character, combine to character
v1_3 <- c(v1,v3)
v1_3

#add vectors
v1 <- v1[1:5]
v1+v2

#adding vectors of different lengths will work - the short vector values will be recycled to 
#align with length of longer vector
v1+v4

#subtract 
v1-v2

#multiply
v1*v2

#logical comparison
v1==v2

#return TRUE/FALSE if value is in vector
3 %in% v1

#sort
v1
sort(v1,decreasing=TRUE)

#sort - character values, descending
v3
sort(v3,decreasing=TRUE)
sort(v3)

#calculations
v_abs<-c(-1,-4,3,7)
abs(v_abs)

```


**Lists**
Ordered collection of objects

Creating list and named lists
```{r}
#create basic list
list1 <- list("dog",375)
list1

#create named list
named_list <- list(pet="dog",number=375)

#create lists with diverse set of objects - character vector, matrix, and a list 
big_list <- list(testing_plain,named_v,list1)

multi_list<- list(c("Sat","Sun"), matrix(c(3,9,6,12,-3,21), nrow = 2), list("dog",375))

#give names to the elements in the list
multi_list<- list("Weekend" = c("Sat","Sun"), "Mults of 3" = matrix(c(3,9,6,12,-3,21), nrow = 2),"Another list"=list("dog",375))

#another option
names(multi_list) <- c("Weekend","Mults of 3","Another list")

```

Indexing lists
- [ ] or $
```{r}
#index list
named_list$pet

#compare single brackets
named_list[2]

#to double brackets
named_list[[2]]

#can also return just names
names(named_list[1])

#indexing in list of list
multi_list[1]
multi_list["Weekend"]
multi_list$Weekend

multi_list[2]

multi_list[3]

multi_list[3][[1]][1]
```

Modifying lists
```{r}

#replace element
multi_list[3][[1]][1] <- "cat"
multi_list[3][[1]][1]

#add new element
multi_list[3][[1]][2] <- "mouse"
multi_list[3][[1]]

multi_list[4] <- "new"
multi_list

#remove the last element.
multi_list[4] <- NULL

#convert to vector
list_a <- list(2,4,6,8)
list_b <- list(1,2,3,4)

#try adding together (uncomment to see error)
# list_a+list_b

vec_a <- unlist(list_a)
vec_b <- unlist(list_b)

vec_a+vec_b

```


