---
title: "Epi R Handbook tour"
output:
  pdf_document: default
  html_notebook: default
---

## 19 Univariate and multivariable regression

https://epirhandbook.com/en/univariate-and-multivariable-regression.html

`pacman` package loads all these libraries and installs them if necessary.
```{r include=F}
# install.packages("pacman")
pacman::p_load(
  rio,          # File import
  here,         # File locator
  tidyverse,    # data management + ggplot2 graphics, 
  stringr,      # manipulate text strings 
  purrr,        # loop over objects in a tidy way
  gtsummary,    # summary statistics and tests 
  broom,        # tidy up results from regressions
  lmtest,       # likelihood-ratio tests
  parameters,   # alternative to tidy up results from regressions
  see          # alternative to visualise forest plots
  )
```

### Univariate

Import data on cases from a simulated Ebola epidemic. 
```{r}
# import the linelist
linelist <- import("linelist_cleaned.rds")

## define variables of interest 
explanatory_vars <- c("gender", "fever", "chills", "cough", "aches", "vomit")

## convert dichotomous variables to 0/1 
linelist <- linelist %>%  
  mutate(across(                                      
    .cols = all_of(c(explanatory_vars, "outcome")),  ## for each column listed and "outcome"
    .fns = ~case_when(                              
      . %in% c("m", "yes", "Death")   ~ 1,           ## recode male, yes and death to 1
      . %in% c("f", "no",  "Recover") ~ 0,           ## female, no and recover to 0
      TRUE                            ~ NA_real_)    ## otherwise set to missing
    )
  )

## add in age_category to the explanatory vars 
explanatory_vars <- c(explanatory_vars, "age_cat")

## drop rows with missing information for variables of interest 
linelist <- linelist %>% 
  drop_na(any_of(c("outcome", explanatory_vars)))

head(linelist)
```



Logistic regression on multiple binomial independent variables:
```{r}
univ_tab <- linelist %>% 
  dplyr::select(explanatory_vars, outcome) %>% ## select variables of interest
  tbl_uvregression(                         ## produce univariate table
    method = glm,                           ## define regression want to run (generalised linear model)
    y = outcome,                            ## define outcome variable
    method.args = list(family = binomial),  ## define what type of glm want to run (logistic)
    exponentiate = TRUE                     ## exponentiate to produce odds ratios (rather than log odds)
  )

## view univariate results table 
univ_tab
```

### Multivariate

No interaction between the explanatory variables (+'s only)
```{r}
mv_reg <- glm(outcome ~ gender + fever + chills + cough + aches + vomit + age_cat, family = "binomial", data = linelist)

summary(mv_reg)
```

Interaction between age and fever:
```{r}
model0 <- glm(outcome ~ gender + age_cat * fever, family = "binomial", data = linelist)
tidy(model0, exponentiate = TRUE, conf.int = TRUE)
```

Using the `plot` function from the [see](https://easystats.github.io/see/) package to visualize odds ratios.
```{r}
# load
pacman::p_load(easystats)

## remove the intercept term from your multivariable results
model0 %>% 
  model_parameters(exponentiate = TRUE) %>% 
  plot()
```

Compare models using a likelihood ratio test:
```{r}
model1 <- glm(outcome ~ age_cat, family = "binomial", data = linelist)
model2 <- glm(outcome ~ age_cat + gender, family = "binomial", data = linelist)

lmtest::lrtest(model1, model2)
```
## 21 Standardised rates

https://epirhandbook.com/en/standardised-rates.html

```{r include=F}
pacman::p_load(
     rio,                 # import/export data
     here,                # locate files
     tidyverse,           # data management and visualization
     stringr,             # cleaning characters and strings
     frailtypack,         # needed for dsr, for frailty models
     dsr,                 # standardise rates
     PHEindicatormethods) # alternative for rate standardisation
```

Import and clean data. These are "fictitious population counts and death counts from country A and country B, by age (in 5 year categories) and sex (female, male)". We combine them all into one data frame.
```{r}
# import demographics for country A directly from Github
A_demo <- import("https://github.com/appliedepi/epirhandbook_eng/raw/master/data/standardization/country_demographics.csv") %>% mutate(Country = "A")

# fix errant "+" in age category
A_demo$age_cat5[[18]] <- 85

# import deaths for country A directly from Github
A_deaths <- import("https://github.com/appliedepi/epirhandbook_eng/raw/master/data/standardization/deaths_countryA.csv")

# import demographics for country B directly from Github
B_demo <- import("https://github.com/appliedepi/epirhandbook_eng/raw/master/data/standardization/country_demographics_2.csv") %>% mutate(Country = "B")

# import deaths for country B directly from Github
B_deaths <- import("https://github.com/appliedepi/epirhandbook_eng/raw/master/data/standardization/deaths_countryB.csv")

# import demographics for country B directly from Github
standard_pop_data <- import("https://github.com/appliedepi/epirhandbook_eng/raw/master/data/standardization/world_standard_population_by_sex.csv")

pop_countries <- A_demo %>%  # begin with country A dataset
     bind_rows(B_demo) %>%        # bind rows, because cols are identically named
     pivot_longer(                       # pivot longer
          cols = c(m, f),                   # columns to combine into one
          names_to = "Sex",                 # name for new column containing the category ("m" or "f") 
          values_to = "Population") %>%     # name for new column containing the numeric values pivoted
     mutate(Sex = recode(Sex,            # re-code values for clarity
          "m" = "Male",
          "f" = "Female"))

deaths_countries <- A_deaths %>%    # begin with country A deaths dataset
     bind_rows(B_deaths) %>%        # bind rows with B dataset, because cols are identically named
     pivot_longer(                  # pivot longer
          cols = c(Male, Female),        # column to transform into one
          names_to = "Sex",              # name for new column containing the category ("m" or "f") 
          values_to = "Deaths") %>%      # name for new column containing the numeric values pivoted
     rename(age_cat5 = AgeCat)      # rename for clarity

country_data <- pop_countries %>% 
     left_join(deaths_countries, by = c("Country", "age_cat5", "Sex"))

country_data <- country_data %>% 
  mutate(
    Country = fct_relevel(Country, "A", "B"),
      
    Sex = fct_relevel(Sex, "Male", "Female"),
        
    age_cat5 = fct_relevel(
      age_cat5,
      "0-4", "5-9", "10-14", "15-19",
      "20-24", "25-29",  "30-34", "35-39",
      "40-44", "45-49", "50-54", "55-59",
      "60-64", "65-69", "70-74",
      "75-79", "80-84", "85")) %>% 
          
  arrange(Country, age_cat5, Sex)

standard_pop_data <- standard_pop_data %>% rename(age_cat5 = AgeGroup)

# Remove specific string from column values
standard_pop_clean <- standard_pop_data %>%
     mutate(
          age_cat5 = str_replace_all(age_cat5, "years", ""),   # remove "year"
          age_cat5 = str_replace_all(age_cat5, "plus", ""),    # remove "plus"
          age_cat5 = str_replace_all(age_cat5, " ", "")) %>%   # remove " " space
     
     rename(pop = WorldStandardPopulation)   # change col name to "pop", as this is expected by dsr package

all_data <- left_join(country_data, standard_pop_clean, by=c("age_cat5", "Sex")) #%>% drop_na()
all_data
```

Direct standardization using [phe_dsr](https://rdrr.io/cran/PHEindicatormethods/man/phe_dsr.html) and [phe_rate](https://rdrr.io/cran/PHEindicatormethods/man/phe_rate.html):
```{r}
# Calculate rates per country directly standardized for age and sex
mortality_ds_rate_phe <- all_data %>%
     group_by(Country) %>%
     PHEindicatormethods::phe_dsr(
          x = Deaths,                 # column with observed number of events
          n = Population,             # column with non-standard pops for each stratum
          stdpop = pop,               # standard populations for each stratum
          stdpoptype = "field")       # either "vector" for a standalone vector or "field" meaning std populations are in the data  
mortality_ds_rate_phe

# crude rate
all_data %>%
     group_by(Country) %>%
     PHEindicatormethods::phe_rate(x = Deaths, n = Population)
```

Indirect standardization. We are "calculating rates for country A using country B as the reference population" using [phr_isr](https://rdrr.io/cran/PHEindicatormethods/man/phe_isr.html):
```{r}
# Create reference population
refpopCountryB <- country_data %>% 
  filter(Country == "B") 

# Calculate rates for country A indirectly standardized by age and sex
mortality_is_rate_phe_A <- country_data %>%
     filter(Country == "A") %>%
     PHEindicatormethods::phe_isr(
          x = Deaths,                 # column with observed number of events
          n = Population,             # column with non-standard pops for each stratum
          x_ref = refpopCountryB$Deaths,  # reference number of deaths for each stratum
          n_ref = refpopCountryB$Population)  # reference population for each stratum

mortality_is_rate_phe_A
```

## 32 Epidemic curves

https://epirhandbook.com/en/ggplot-tips.html

```{r include=FALSE}
pacman::p_load(
  rio,          # file import/export
  here,         # relative filepaths 
  lubridate,    # working with dates/epiweeks
  aweek,        # alternative package for working with dates/epiweeks
  incidence2,   # epicurves of linelist data
  i2extras,     # supplement to incidence2
  stringr,      # search and manipulate character strings
  forcats,      # working with factors
  RColorBrewer, # Color palettes from colorbrewer2.org
  tidyverse     # data management + ggplot2 graphics
) 
```

Same data on cases from a simulated Ebola epidemic from the regression section:
```{r}
head(linelist)
```

```{r}
# create the incidence object, aggregating cases by day
epi_day <- incidence(       # create incidence object
  x = linelist,             # dataset
  date_index = date_onset,  # date column
  interval = "day"          # date grouping interval
  )

epi_day
# plot the incidence object
plot(epi_day)

```
```{r}
# Weekly (Monday week by default)
epi_wk      <- incidence(linelist, date_onset, interval = "Monday week")
plot(epi_wk)

# Monthly
epi_month   <- incidence(linelist, date_onset, interval = "month")
plot(epi_month)
```




