---
title: "PHW251 Problem Set 7"
author: "Tin Ho"
date: "2022-11-11"
output: pdf_document
---


```{r, include=F}
# load libraries and suppress output w/ include=F
library(tidyverse)
```

## Part 1

For part 1 of this problem set we will work with motor vehicle crash data from New York City. You can read more about this [publicly available data set on their website](https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95).

The data file is called "Motor_Vehicle_Collisions_Crashes.csv". We want you to perform the following:

1. Rename the column names to lower-case and replace spaces with an underscore.
2. Select only:
   + crash_date
   + number_of_persons_injured
   + contributing_factor_vehicle_1
   + vehicle_type_code_1
3. Drop all rows that contain an NA value.
4. Make the values in the vehicle_type_code_1 variable all lowercase and replace the spaces with a dash.
5. Filter the data for vehicles that have a count of at least 500 (appear in the data set 500 times or more)
   + Hints: group_by(), mutate(), n(), filter()
6. Calculate the percentage of accidents by vehicle type
7. Which vehicle group accounted for 1.55% (0.0155) of the accidents?

We have grouped the questions below to push you to perform commands with less code. As you're building your code we recommend going line by line to test, then combining to perform multiple steps in one command.

\newpage

### Questions 1-3

```{r, include=F}
# load your data
df_motor <- read_csv("data/Motor_Vehicle_Collisions_Crashes.csv")

```


```{r}
# YOUR CODE HERE
# Q1,2,3:
df_motor_q1 = df_motor %>%
  rename_all( . %>% tolower %>% gsub( " ", "_", . )  
            ) %>%
  select( crash_date, 
          number_of_persons_injured,
          contributing_factor_vehicle_1,
          vehicle_type_code_1 ) %>%
  drop_na()

str(df_motor_q1)

```

\newpage

### Questions 4-5

```{r}
# YOUR CODE HERE

# Q4.... Make the values in the vehicle_type_code_1 variable 
# all lowercase and replace the spaces with a dash.
df_motor_q1$vehicle_type_code_1   = 
  str_replace_all( df_motor_q1$vehicle_type_code_1, " ", "-"  ) 

df_motor_q1$vehicle_type_code_1 = 
  tolower( df_motor_q1$vehicle_type_code_1 ) 

#5. Filter the data for vehicles that have a count of at least 500 
#   (appear in the data set 500 times or more)
#   + Hints: group_by(), mutate(), n(), filter()

df_motor_q5 = df_motor_q1 %>%
  group_by( vehicle_type_code_1 ) %>%
  count( vehicle_type_code_1 ) %>%
  filter( n >= 500 )

df_motor_q5 %>% head


```

\newpage

### Question 6

```{r}
# YOUR CODE HERE
#6. Calculate the percentage of accidents by vehicle type

## so this is % after having filter for n >= 500, 
## NOT % as calculated from the full data set.
## keeping it this way cuz the questions were sequentially numbered this way.

total_cases = sum(df_motor_q5$n)

df_motor_q6 = df_motor_q5 %>% 
  mutate( pct = round( n/total_cases * 100, 2)  )

df_motor_q6 %>% head

#7. Which vehicle group accounted for 1.55% (0.0155) of the accidents?

df_motor_q6 %>% 
  filter( pct == 1.55 )

```

### Question 7

WRITE YOUR ANSWER HERE

_bus_

\newpage

# Part 2

For this part we will work with four tables that are relational to each other. The following keys link the tables together:

* patient_id: patients, schedule
* visit_id: schedule, visits
* doctor_id: visits, doctors

```{r, echo=F, include=F, results="markup"}
patients <- read_csv("data/patients.csv")
visits <- read_csv("data/visits.csv")
doctors <- read_csv("data/doctors.csv")
schedule <- read_csv("data/schedule.csv")

patients  %>% head()
visits    %>% head()
schedule  %>% head()

```

### Question 8

You've been asked to collect information on patients who are actually on the schedule. To start this task, you need to join the patient data to the schedule data, since we only want to keep the observations that are present in both the patient data AND the schedule data.

Which kind of join do you use?

WRITE YOUR ANSWER HERE

_inner join_

How many observations do you see in your joined data set? Notice that some patients have multiple visits.

```{r}
# YOUR CODE HERE

patient_schedule = 
  inner_join( 
              x = patients, 
              y = schedule,
              by = "patient_id"
              )

# right_join => 124 rec
# inner_join => 124 rec # still same result cuz schedule always have valid patient_id


#str( patient_schedule ) # 124 rec
num_obs =  patient_schedule %>% tally()
as.numeric( num_obs )

```

WRITE YOUR ANSWER HERE

_124_

\newpage

### Question 9

In the visits data, we have a variable called "follow_up" where Y means
a follow-up is needed and N means a follow-up is not needed. How many patients require a follow-up? 
You will want to first make a join and then subset. Start with the data frame created in the previous question.

```{r}
# YOUR CODE HERE

patient_schedule_visit =
  inner_join( 
              x = patient_schedule,
              y = visits,
              by = "visit_id" )  

# full_join  => 124 rec, some has NA in follow_up
# inner_join => 109 rec, no NA
  
follow_up = 
  patient_schedule_visit %>%
    filter( follow_up == "Y" ) 

num_follow_up = follow_up %>% tally()
as.numeric( num_follow_up )

```

Which join did you use?

WRITE YOUR ANSWER HERE

_inner join_

How many patients need a follow-up?

WRITE YOUR ANSWER HERE

_27_

\newpage

### Question 10

Which doctors do these patients need follow-up with? Print out each
doctor's name.

```{r}
# YOUR CODE HERE

doc_follow = 
  inner_join( 
              x = doctors,
              y = follow_up,
              by = "doctor_id"  )

doc_follow %>% pull( doctor ) %>% unique()


```

Which join did you use?

WRITE YOUR ANSWER HERE

_inner join_





