---
title: "Toolkit - Data Frames"
author: "Lauren Nelson"
date: "9/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Data frames
* Creating data frames
  + From vectors
  + From a built in dataset
* Creating new rows and columns
* Indexing, extracting, reassigning 
* Exploring data frames
* Data frames versus tibbles

### Creating from vectors
Data frames can be created from vectors using the data.frame() function. Vectors must be of the same length. 
```{r}

#defaults for data.frame() for reference
# data.frame(..., 
#   row.names = NULL, 
#   check.rows = FALSE,
#   check.names = TRUE, 
#   fix.empty.names = TRUE,
#   stringsAsFactors = default.stringsAsFactors())

#create data frame without specifying options (will use the defaults specified in the 
#commented out section above)
testing_plain <- data.frame(
  county = c("Alameda","Contra Costa","Marin"),
  total_tests = c(500,745,832),
  pos_tests = c(43, 32, 30)
)

#look at the structure of the data frame
#note that county is reading in as a 'factor', which will make it challenging to add to or 
#manipulate the data frame
#(factors will be covered in more detail in the next few weeks)
str(testing_plain)

#set stringsAsFactors=FALSE to allow character values to read in as character, rather than 
#factor
testing <- data.frame(
  county = c("Alameda","Contra Costa","Marin"),
  total_tests = c(500,745,832),
  pos_tests = c(43, 32, 30),
  stringsAsFactors = FALSE
)

#check the structure to make sure column types are correct
str(testing)

#print to console
testing

#row names: create data frame which utilizes the row.names option and set the row name to 
testing_row_name <- data.frame(
  county = c("Alameda","Contra Costa","Marin"),
  total_tests = c(500,745,832),
  pos_tests = c(43, 32, 30),
  row.names = "county"
)

#print to console - notice that there are no longer row numbers 1-3, they have been 
#replaced by county (and there is no longer a column labeled 'county')
testing_row_name
```

### Indexing,extracting, reassigning
* [[]] or $ to create a vector
* [] to create a new data frame
```{r}
#using [ ], [[ ]], and $

#output or create a vector of the values in a column
#note: there are multiple ways to do this - all 3 lines below will return the same vector
testing[["county"]]
testing[[1]]
testing$county

#assign the vector to an object
county_vector <- testing[["county"]]

#output a dataframe of the values in a column
county_df <- testing["county"]
county_df2 <- testing[1]

#subset rows
#using the row number - can be 1:2 or c(1,2)
testing[1:2,]  
#based on a specified data elements
testing[which(testing$total_tests>500),]

#subset columns based on column name or number
testing[,c("total_tests","pos_tests")]
testing[,c(2,3)]

#subset rows and columns
testing[c(1,2),c("total_tests","pos_tests")]

#Reassign a value of a cell
testing[2,1] <- "Yolo"

testing

#Reassign column values
testing$pos_tests <- c(52,61,89)

```


### Adding rows
```{r}

#add a row using rbind()
#create list that includes an element for each column - using a list is important since there 
#are data of different types in the data frame
add_county <- list("San Diego", 1200, 73)

#use row bind - will combine in the order given
testing <- rbind(testing,add_county)

#row bind can also be used to combine two compatible data frames 
#(compatible meaning they have the same columns and data types)
testing2 <- data.frame(
  county = c("Imperial","Riverside"),
  total_tests = c(410,760),
  pos_tests = c(84,57),
  stringsAsFactors = FALSE
)

#combine the two data frames
#note: calling this new data frame 'testing' will save over our previous data 
#frame named 'testing'
testing <- rbind(testing,testing2)

```


### Adding columns
```{R}
#add a column
#create a new column called 'id'
testing$id <- 1:6

#another way to do this is to create a vector of column values
id <- 1:6

#and add in using column bind
testing_id2 <- cbind(testing,id)

#adding a column based on other columns
testing$pct_pos <- round(100*testing$pos_tests/testing$total_tests,1)

```

### Restructuring
```{r}
#reorder columns
testing <- testing[,c(4,1,2,3,5)]

#reorder rows based on values in a specified column
testing <- testing[order(testing$county),]

#create a new df without the county column
#option 1: list columns by order or name
testing_nocounty <- testing[,c(1,3,4,5)]

#option 2: use the "-" to indicate which columns to exclude
testing_nocounty2 <- testing[,c(-2)]

#testing_nocounty will be the same as testing_nocounty2

```

### Data frames vs. tibbles 
Tibbles are a type of dataframe that are used within the tidyverse. They were essentially created to be a bit more flexible and user friendly. Key differences: 

* No row names
* Column names can have spaces, start with characters, as long as within back-ticks
* When a tibble is printed to the console, it will automatically only include 10 lines, and will also include the data types
* Can handle a vector of length 1 (will assign value to all rows)
* Can create columns by calculating on others
```{r}
#try creating a tibble version similar to the data frame created above
testing_tib <- tibble(
  county = c("Alameda","Contra Costa","Marin"),
  state = "CA",
  total_tests = c(500,745,832),
  pos_tests = c(43, 32, 30),
  `Percent Positive Tests` = round(100*(pos_tests/total_tests),1)
)

#print to console
testing_tib

#compare tibble...
class(testing_tib)
typeof(testing_tib)

#...to data frame
class(testing)
typeof(testing)

```


### Creating from built in data frames
There are many built in data frames in RStudio. Some are available through base R, others can be accessed through packages. These are great options for testing out new skills, or creating reproducible examples.
```{r}
#generate list of available data in RStudio (uncomment to run)
# data()

#example of a package that includes more dataset options
library(nycflights13)

#this is a good dataset because it includes >300k rows and has both numeric and character #data
#running this line will add "flights" object to the environment pane on the right. 
#it may show up as a <promise>, in which case click on it in the environment pane and 
#the whole dataset will load
data("flights")

#explore flights
#structure
str(flights)

#number of rows and colums
nrow(flights)
ncol(flights)

#type/class of object
class(flights)
typeof(flights)

#summary - mean/median/quartiles
summary(flights$distance)

#unique values
unique(flights$origin)

#frequency
table(flights$carrier)

#cross tab (create as object to make it easier to view)
cross_tab <- data.frame(table(flights$origin, flights$dest))

#add column names to cross tab table
colnames(cross_tab) <- c("origin","dest","freq")

#limit to only pairs with frequency >0
cross_tab <- cross_tab[which(cross_tab$freq>0),]

#plot (use Help for more options)
#specify type = "p" to plot points
plot(flights$dep_time, flights$dep_delay, type = "p")

#specify type = "b" to plot both points and line
plot(flights$month,flights$dep_delay, type = "b")

```


