---
title: "Toolkit - Importing & Exporting"
author: "Lauren Nelson"
date: "9/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(readxl)
```

## Importing & exporting data
* Importing data 
  + Example .csv import
  + Example .xslx import
* Exporting data


### Importing a .csv file

**readr package**

Why readr vs base R (read.csv())?

* The two are fairly similar; many options to specify how your data imports
* Ultimately, readr is easier to use, control, and reproduce
* Part of the tidyverse, so works well with other packages that we will use

[readr::read_csv](https://readr.tidyverse.org/reference/read_delim.html)

Arguments 

* file	- 
  + when using datahub or rstudio cloud, easiest to upload a file to the cloud for import into your session 
  + can also download files starting with http://, https://, ftp://, or ftps:// straight from the internet
* col_names	- 
  + TRUE - first row of the input will be used as the column names
  + FALSE - column names will be generated automatically: X1, X2, X3 etc.
  + character vector of column names - values will be used as the names of the columns
* col_types	
  + NULL - column types imputed from first 1000 rows
  + cols() specification - must contain one column specification for each column (full list [here](https://readr.tidyverse.org/reference/cols.html))
  + cols_only() - specify type for subset of columns to import
  + string - a compact string representation where each character represents one column (c = character, i = integer, 
  n = number, d = double, l = logical, f = factor, D = date, T = date time, t = time, ? = guess, or _/- to skip the column)
* na	- character vector of strings to interpret as missing values
* skip - number of lines to skip before reading data.
* n_max	- maximum number of records to read


```{r, results="hide", warning=FALSE, message=FALSE}

# read_csv(file,
#          col_names = TRUE,
#          col_types = NULL,
#          locale = default_locale(),
#          na = c("", "NA"),
#          quoted_na = TRUE,
#          quote = "\"",
#          comment = "", 
#          trim_ws = TRUE, 
#          skip = 0,
#          n_max = Inf, 
#          guess_max = min(1000, n_max),
#          progress = show_progress(), 
#          skip_empty_rows = TRUE)

#import a file via the web
file_path_csv <- "https://data.chhs.ca.gov/dataset/8eb8839f-52a1-410b-8c4a-5b1c1678bbc2/resource/21fec3a1-ae82-49f4-ae97-fd8c78ca22ee/download/retail-availability-of-electronic-smoking-devices-by-county.csv"

#read in file with no options changed
esd <- read_csv(
  file_path_csv
)

#specify column names
esd_colnames <- read_csv(
  file_path_csv,
  col_names = c("county","yr","pct","ci_l","ci_u")
)

#notice that when specifying column names, any column names in the csv file are now read 
#in as row 1, so use skip to change the first read in to 1 rather than the default of 0
esd_colnames <- read_csv(
  file_path_csv,
  col_names = c("county","yr","pct","ci_l","ci_u"),
  skip = 1
)

#look at data structure
str(esd_colnames)

#notice that the percentage and confidence interval columns are reading in as character
#specify column types using cols(), cols_only(), or the compact string method
#the commented out lines offer different methods; using cols_only() allows to select
#a subset of columns to include
esd_coltypes <- read_csv(
  file_path_csv,
  col_names = c("county","yr","pct","ci_l","ci_u"),
  # col_types = cols(col_character(), col_double(), col_number(), col_number(), col_number()),
  # col_types = cols_only(county = col_character(), yr = col_double(), pct = col_number()), 
  col_types = "cdnnn",
  skip = 1
)

str(esd_coltypes)

#fix the warnings due to unexpected values by adding * and n/a to the list of values set 
#to missing
esd_coltypes <- read_csv(
  file_path_csv,
  col_names = c("county","yr","pct","ci_l","ci_u"),
  col_types = "cdnnn",
  na = c("", "NA","*","n/a"),
  skip = 1
)

str(esd_coltypes)

#save a clean version of the data file
esd_clean <- esd_coltypes

```

## Import from excel

**readxl package**
Additional/different arguments than readr::read_csv

* sheet	-  a string (the name of a sheet), or an integer (the position of the sheet). Ignored if the sheet is specified via range. If neither argument specifies the sheet, defaults to the first sheet.
* range	- A cell range to read from, as described in cell-specification. 
* col_types	- 
  + NULL to guess all from the spreadsheet
  + character vector containing one entry per column from these options: "skip", "guess", "logical", "numeric", "date", "text" or "list". 
```{r, results="hide", warning=FALSE, message=FALSE}
#import from folder within RStudio
#first, must upload desired file 

#see working directory
getwd()

#use working directory to establish path to file (assuming uploaded to RStudio cloud)
file_path_excel <- "/home/rstudio/toolkits/data/tb_in_ca.xlsx"

#read in with no extra arguments specified
tb <- read_excel(
  file_path_excel
)

#structure of table
str(tb)

#utilize arguments to specify column types
tb_cleaner <- read_excel(
  file_path_excel,
  col_types = c("numeric","text","text","numeric","numeric","skip") #specify column types
)

#clear warnings due to values = .; set to missing using the na argument
tb_cleanest <- read_excel(
  file_path_excel,
  col_types = c("numeric","text","text","numeric","numeric","skip"), #specify column types
  na = c(".") #specify values to be considered missing
)

#confirm structure of table is what was desired
str(tb_cleanest)


```

### Exporting data
This is a lot simpler than importing data. Before exporting consider:

* Who will be using the dataset/table?
* Will more analysis need to be done, or are you exporting for someone to look at in excel?
```{r, results="hide", warning=FALSE, message=FALSE}

#export csv - good if needing flexibility for other analytic programs
readr::write_csv(esd_clean,"/home/rstudio/toolkits/data/esd_clean.csv")

#export to excel - good if others want to look at it in excel (not R)
openxlsx::write.xlsx(esd_clean,"/home/rstudio/toolkits/data/esd_clean.xslx")

#export RDS - good if you will want to pick it back up in R
saveRDS(esd_clean,"/home/rstudio/toolkits/data/esd_clean_saved.RDS")

```

### Other file formats
Check out [haven](https://haven.tidyverse.org/) for importing/exporting SAS, SPSS, Stata